---
title: "Assignment 3"
output: html_notebook
---

1. (1p) Download the dataset college.csv and explore its overall structure. Get a summary statistics of each
variable. Answer the following questions:
• How many observation do you have in the data?
• How many categorical and numeric variables you have in your data?

```{r}
clg_df = read.csv("College.csv")
str(clg_df)
```

There are 777 observations, 2 categorical variables and 17 numeric variables.

• Is there any missing value?
```{r}
sum(is.na(clg_df))
```
There appear to be no missing values in the dataset

2. (1pt) Remove the first column (the name of the college)
```{r}
clg_df = clg_df[,-1]
```

3. (2 pts) Which variables are associated with “Apps”. Use “pairs”, “cor”, and side by side box plot with
t.test to answer this question.

```{r}
clg_df = as.data.frame(unclass(clg_df), stringsAsFactors = TRUE)
attach(clg_df)
```
Converting strings to factors and attaching the dataframe
```{r}
iqr = 3264 - 776
iqr_index = (Apps > 776 - 1.5*iqr & Apps < 3264 + 1.5*iqr)
Apps_so = Apps[iqr_index] #so means sans outliers
Private_so = Private[iqr_index]
plot(Apps_so~Private_so)
t.test(Apps_so~Private_so)
```
```{r}
pairs(clg_df[2:6])
pairs(clg_df[c(2,7:9)])
pairs(clg_df[c(2,10:12)])
pairs(clg_df[c(2,13:15)])
pairs(clg_df[c(2,16:18)])
```

```{r}
cor(Apps, clg_df[-c(1, 2)])
```

From the above tests I've concluded that private, Accept, Enroll, F.Undergrad, and Private are associated with Apps

4. (1pt) plot the histogram of the number of applications “Apps” variable. Explain what the histogram shows?
```{r}
hist(Apps)
```

Judging from the shape this historgram is right-skewed (possibly j-shaped if you see the small hump between 45k and 50k) with a mean between 0 and 10000

5. (1pt) replace the “Top10perc” variable with a factor variable “Elite” with two levels: “Yes” if 50% or more of new students coming from the top 10% of their high school class (that is, if Top10Perc>=50) , and “No” if less than 50% of new students are coming from the top10% of their high school class.

```{r}
clg_df$Elite = factor(ifelse(clg_df$Top10perc >= 50, "Yes", "No"))
clg_df = clg_df[-5]
```

6. (1pt) is the Elite variable you created above associated with the “Apps” variable? Don’t just say yes or no, you should explain your answer by plots and test statistics.

```{r}
Elite_so = clg_df$Elite[iqr_index]
plot(Apps_so~Elite_so)
t.test(Apps_so~Elite_so)
```

Based on the side-by-side box plot and the p-value = 0.01% < 5% it follows that Elite is associated with Apps

7. (1pt) normalize all numerical features (except the “Apps” variable) using z-score standardization using “scale” function
```{r}

```

