---
title: "Assignment 3"
output: html_notebook
---

1. (1p) Download the dataset college.csv and explore its overall structure. Get a summary statistics of each
variable. Answer the following questions:
• How many observation do you have in the data?
• How many categorical and numeric variables you have in your data?

```{r}
clg_df = read.csv("College.csv")
str(clg_df)
```

There are 777 observations, 2 categorical variables and 17 numeric variables.

• Is there any missing value?
```{r}
sum(is.na(clg_df))
```
There appear to be no missing values in the dataset

2. (1pt) Remove the first column (the name of the college)
```{r}
clg_df = clg_df[,-1]
```

3. (2 pts) Which variables are associated with “Apps”. Use “pairs”, “cor”, and side by side box plot with
t.test to answer this question.

```{r}
clg_df = as.data.frame(unclass(clg_df), stringsAsFactors = TRUE)
attach(clg_df)
```
Converting strings to factors and attaching the dataframe
```{r}
iqr = 3264 - 776
iqr_index = (Apps > 776 - 1.5*iqr & Apps < 3264 + 1.5*iqr)
Apps_so = Apps[iqr_index] #so means sans outliers
Private_so = Private[iqr_index]
plot(Apps_so~Private_so)
t.test(Apps_so~Private_so)
```
```{r}
pairs(clg_df[2:6])
pairs(clg_df[c(2,7:9)])
pairs(clg_df[c(2,10:12)])
pairs(clg_df[c(2,13:15)])
pairs(clg_df[c(2,16:18)])
```

```{r}
cor(Apps, clg_df[-c(1, 2)])
```

From the above tests I've concluded that private, Accept, Enroll, F.Undergrad, and Private are associated with Apps